A multi-species benchmark for training and validating mass spectrometry proteomics machine learning models

### 2024-08-29 doc/nine-species/response.tex

(2) Why did the author only consider the FDR at PSM level? I think the FDR at protein level is commonly used now.

\response{We agree that, in many settings, controlling the FDR at the protein level is appropriate, especially when a study is aimed at characterizing the protein content of a given set of biological samples.
  However, for constructing a benchmark of PSMs used for training and validating \textit{de novo} sequencing methods, we believe that protein level FDR control is not as relevant.
  In particular, even if we control the FDR at the protein level, the benchmark needs to report a set of PSMs rather than proteins, and it is not clear how to go from a set of detected proteins to a corresponding collection of confident PSMs.
  Consider, for example, the widely used ``picked protein'' method for protein-level FDR control \cite{savitski2015scalable}.
  In the Savitski \textit{et al.}\ (2015) description of this method, each protein is assigned a score equal to the sum of the search engine scores assigned to its peptides.
  With such an approach, a protein can be detected with high confidence even if one or more of its associated peptides has low scores, as long as some of the peptides have very high scores.
  Thus, if we include all PSMs associated with a detected peptide in our benchmark, we risk contaminating the benchmark with incorrect PSMs.}

(3) The annotation in MGF is adding the peptide sequence. Is it possible to provide the annotated product ions? For example, which peak is annotated as b5 ion or y3 ion. With these annotations at ion level, this dataset will be more useful.

\response{Yes, we have added to the benchmark, alongside each MGF file a corresponding mzSpecLib \cite{klein2024proteomics} file that provides this annotation.}

### 2024-08-29 doc/nine-species/nine-species.tex
Update response

\author[1]{Wout Bittremieux}
\author[2,3]{William Stafford Noble\footnote{Correspondence: william-noble@uw.edu}}

@@

The dataset contains files resulting from various steps in the generation of the benchmark:
\begin{itemize}
\item Spectrum files in MGF format, produced by msconvert.
\item Reference proteome files in FASTA format, downloaded from Uniprot.
\item Search results files for both targets and decoys, in tab-delimited format, produced by Tide.
\item PSM-level Percolator results files for targets, in tab-delimited format.
\item Annotated MGF files for both versions of the benchmark (main and balanced).
\end{itemize}
Also included are log files for the steps of the analysis pipeline carried out using Crux \cite{park2008rapid} (Tide indexing, Tide search, and Percolator).
The data is available at \url{https://doi.org/10.5281/zenodo.12819175}.

@@

\begin{figure}
  \centering
  \begin{tabular}{lll}
    \includegraphics[height=4in]{../../results/wnoble/2024-07-17validations/nine-species-fdrs.pdf} &
    \includegraphics[height=4in]{../../results/wnoble/2024-07-17validations/percent-id.pdf}  &
    \includegraphics[height=4in]{../../results/wnoble/2024-07-17validations/match-by.pdf}  \\
    a & b & c \\
  \end{tabular}
  \caption{\textbf{Validation of the benchmarks.}
    \textbf{a.} Each series indicates, for a given species, the number of accepted PSMs as a function of PSM-level FDR.
    \textbf{b.} The bar plot indicates the proportion of spectra that were accepted at 1\% PSM-level FDR per species.
    \textbf{c.} Each histogram shows, for one species, the distribution of the proportion of total ion current that is matched by b- or y-ions per accepted PSM, using a matching tolerance of 0.05~\textit{m/z}.}
  \label{fig:fdr}
\end{figure}

>>>>

\author[1]{Wout Bittremieux}
\author[2]{Bo Wen}
\author[2,3]{William Stafford Noble\footnote{Correspondence: william-noble@uw.edu}}

@@

The dataset contains files resulting from various steps in the generation of the benchmark:
\begin{itemize}
\item Spectrum files in MGF format, produced by msconvert.
\item Reference proteome files in FASTA format, downloaded from Uniprot.
\item Search results files for both targets and decoys, in tab-delimited format, produced by Tide.
\item PSM-level Percolator results files for targets, in tab-delimited format.
`\item Annotated MGF and corresponding mzSpecLib \cite{klein2024proteomics} files for both versions of the benchmark (main and balanced).
\end{itemize}
Also included are log files for the steps of the analysis pipeline carried out using Crux \cite{park2008rapid} (Tide indexing, Tide search, and Percolator).
The data is available at \url{https://doi.org/10.5281/zenodo.12819175}.

@@

\begin{figure}
  \centering
  \begin{tabular}{lll}
    \includegraphics[height=4in]{nine-species-fdrs.pdf} &
    \includegraphics[height=4in]{percent-id.pdf}  &
    \includegraphics[height=4in]{match-by.pdf}  \\
    a & b & c \\
  \end{tabular}
  \caption{\textbf{Validation of the benchmarks.}
    \textbf{a.} Each series indicates, for a given species, the number of accepted PSMs as a function of PSM-level FDR.
    \textbf{b.} The bar plot indicates the proportion of spectra that were accepted at 1\% PSM-level FDR per species.
    \textbf{c.} Each histogram shows, for one species, the distribution of the proportion of total ion current that is matched by b- or y-ions per accepted PSM, using a matching tolerance of 0.05~\textit{m/z}.}
  \label{fig:fdr}
\end{figure}
